---
title: "R day2"
author: "Kejin Qian"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
here::here("data", "gapminder5.csv")

gapminder <- read.csv(here::here("data","gapminder5.csv"))
```


##Converting factors to characters
```{r}
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
```

## R for loop 
```{r}
#create a vector of values that you want to repeat the function for
#this is the number of times you are going to loop with
obs <- 1 : nrow(gapminder)
#initialize the for loop with 'for (i in vector)'
for (i in obs){#the function to repeat is enclosed in braces {}
  gapminder[i, "gdp"] <- gapminder[i, 'pop']* gapminder[i,'gdpPercap']}
```

## Creating logged variables with a for loop
```{r}
#note you can put as many functions as you want in the same loop
obs <- 1 : nrow(gapminder)
for (i in obs){gapminder[i, 'logGdp'] <- log(gapminder[i, 'gdp'])
gapminder[i, 'logPop']<- log(gapminder[i,'pop'])}
```

## A way to avoid loop
```{r}
gapminder$vec_log_gdp <- log(gapminder$gdp)
all(gapminder$vec_log_gdp == gapminder$logGdp)
```

## Has life expectancy increased over year
```{r}
years <- unique(gapminder$year)
for (i in years){
  mean_le <- mean(gapminder$lifeExp[gapminder$year == i], na.rm = T)
  print(paste0(i, ": ", mean_le))
}
```

## Mean life expectancy by continent
```{r}
continent <- unique(gapminder$continent)
for (i in continent){
  mean_lifeep <- mean(gapminder$lifeExp[gapminder$continent == i],na.rm =T)
  print(paste0(i, ": ", mean_lifeep))
}
```

## Nested loop
```{r}
for (i in continent) {
  print(paste0("Continent: ", i))
  for (j in years) {
    mean_le <- mean(gapminder$lifeExp[gapminder$continent == i & 
                                          gapminder$year == j], 
                      na.rm = T)
    print(paste0(j, ": ", mean_le))
  }
}
```

## Nested Loop ex
```{r}
for (i in continent) {
  print(paste0("Continent: ", i))
  for (j in years) {
    std_le <- sd(gapminder$lifeExp[gapminder$continent == i & gapminder$year == j],
                 na.rm = T)
    print(paste0(j, ": ", std_le))
  }
}
```

## Apply
```{r}
vars <- gapminder[,c('lifeExp','pop','gdpPercap')]
apply(vars, 2, mean, na.rm =T)
```

## Compare Apply with for loop
```{r}
print(names(vars))
#"lifeExp"   "pop"       "gdpPercap"
for (i in vars) {print(mean(i))}
#mean of each column
```

## The while loop
```{r}
i <- 1952
while (i < 1987) {sd_le <- sd(gapminder$lifeExp[gapminder$year == i])
    print(paste0(i, ": ", sd_le)
    )
    i <- i + 5
    }
```
```{r}
i <- 1987
while (i < 2003) {sd_le <- sd(gapminder$lifeExp[gapminder$year == i])
    print(paste0(i, ": ", sd_le)
    )
    i <- i + 5
    }
```

## IF/ELSE Statement
```{r}
random_year <- sample(years, 1)
if (random_year > 1980) {print(random_year)} else {
  print('sorry, random year is less than 1980')}
```

## Put for and IF/ELSE together
```{r}
threshold <- 70

for (i in unique(gapminder$continent)) {
   tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
   
   if (tmp < threshold) {
       print(paste("Mean Life Expectancy in", i, "is less than", threshold))
   } else {
       print(paste("Mean Life Expectancy in", i, "is greater than", threshold))
   }
}
```
```{r}
yr <- 1987
for (i in years){
  if (i >= 1987) {
  tmp <- mean(gapminder$pop[gapminder$year == i])
  print(paste("Mean population in", i, "is", tmp))
  }else{print(paste("Error Message, the year is less than", yr))}
  
}
```

## Writing Functions
```{r}
get_values <-
  function(df, variable = "continent") {
    vals <- unique(df[[variable]])  
    print(paste0(variable, ": ", vals))
  }
get_values(gapminder,'country')
```

## Write Functions
```{r}
report_mean_sd <- 
function(df, variable, country) {
    var <- df[[variable]][df$country == country]
    m_le <- mean(var)
    sd_le <- sd(var)
    cat("Country:", country, 
        "\nMean Life Expectancy:", m_le,
        "\nSD Life Expectancy:", sd_le)
  }

report_mean_sd(gapminder, "lifeExp", "Bulgaria")
```

## Write Function Ex
```{r}
report_stat <-
function(df, variable, continent) {
  var <- df[[variable]][df$continent == continent]
  max = max(var)
  min = min(var)
  me = mean(var)
  mdm = median(var)
  cat("Continent:", continent,
      "\nMean Life Expectancy:", me, "\nMedium Life Expectancy:", mdm,
      "\nMax Life Expectancy:", max, "\nMin Life Expectancy:", min)
}
report_stat(gapminder,'lifeExp','Asia')
```

## Day2 Exercise
## IF Statement
```{r}
yrs = unique(gapminder$year)
#try year 2002
if (any(yrs == 2002)) {print("There is information of year 2002")} else {
  print("Sorry, no information for year 2002")}
#try year 2012
if (any(yrs == 2012)) {print("There is information of year 2002")} else {
  print("Sorry, no information for year 2002")}
```

## LOOP AND IF STATEMENT
```{r}
#Write a script that finds the mean life expectancy by country for countries whose #population is below the mean for the dataset
mean_pop = mean(gapminder$pop)
df = gapminder[['lifeExp']][gapminder$pop > mean_pop]
mean(df)
```

```{r}
#Write a script that loops through the gapminder data by continent and prints out #whether the mean life expectancy is smaller than 50, between 50 and 70, or greater #than 70.
cont = unique(gapminder$continent)
for (i in cont) {mean_le = round(mean(gapminder[['lifeExp']][gapminder$continent == i]),2)
print(paste("The mean life expectancy in", i, "is", mean_le)) 
if (mean_le < 50) {print("Mean life expectancy is smaller than 50")} else if (mean_le > 70) {print("Mean life expectancy is larger than 70")} else {print("Mean life expectancy is between 50 and 70")}}
```

## Write Functions
```{r}
#Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset. Hint: Use mode() or class() to get the class of the data in each column. Remember that names() or colnames() returns the name of the columns in a dataset.
dataset_summary <-
  function(df){
    colname = names(df)
    print(colname)
    for (i in df){print(class(i))}
  }
dataset_summary(gapminder)
```

```{r}
#Create a function that given a vector will print the mean and the standard deviation of a vector, it will optionally also print the median. Hint: include an argument that takes a boolean (TRUE/FALSE) operator and then include an if statement.
mean_sd_median <-
  function(vec,med = FALSE)
  {
    print(paste("mean is", mean(vec))) 
    print(paste("sd is", sd(vec)))
    if (med == TRUE) 
      {print(paste("median is", median(vec)))}}
mean_sd_median(gapminder$pop, TRUE)
```

## Relationship
```{r}
yr = unique(gapminder$year)
viz_lm <-
  function(df, dv, iv, year) {
    dat <- df[df[["year"]] == year, ]
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x)
    plot(y ~ x, main = year,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'blue')
  }
for (i in yr) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)}
```
##Relationship ex
```{r}
n = 1 : nrow(gapminder)
for (i in n) {
if (gapminder[i,'country'] == 'Canada'| gapminder[i,'country'] == 'United States'| gapminder[i,'country'] == 'Mexico'){gapminder[i,'continent_new'] = 'North America'} else if (gapminder[i,'continent'] == "Americas"){gapminder[i,'continent_new'] = 'South America'} else {gapminder[i,'continent_new'] = gapminder[i,'continent']}}

```
## CONTINUE
```{r}

ctnew = unique(gapminder$continent_new)
viz_lm <-
  function(df, dv, iv, continent) {
    dat <- df[df[["continent_new"]] == continent, ]
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x)
    plot(y ~ x, main = continent,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'blue')
  }
for (i in ctnew) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)}
```


